---
title: "Survey on Chat/Q&A Platforms"
output: github_document
params:
  username:
    label: "LimeSurvey username"
    value: ""
    input: text  
  password:
    label: "LimeSurvey password"
    value: ""
    input: text
---

```{r setup, include=FALSE}
# library(devtools)
# install_github("cloudyr/limer")
library(pals)
library(limer)
library(dplyr)
library(forcats)
library(ggplot2)
library(patchwork)
library(readr)
# Setup API details and login to LimeSurvey
options(lime_api = 'https://rfoundation.limequery.com/admin/remotecontrol')
if (interactive()){
    options(lime_username = rstudioapi::askForPassword(prompt = "username"))
    options(lime_password = rstudioapi::askForPassword())
} else{
    options(lime_username = params$username)
    options(lime_password = params$password)
}
get_session_key()
# Setup knitr
knitr::opts_chunk$set(echo = FALSE)
# fill colours
scale_fill_discrete <- function(...) {
   scale_fill_manual(..., values = kelly(7)[-1])
} 
```

```{r data, include = FALSE}
responses <- get_responses(129653)
```

## Background

Following up on the idea of a mentorship forum for R core contributors (discussed in [this issue](https://github.com/forwards/rcontribution/issues/5) and at our [September 2020 meeting]( https://github.com/forwards/rcontribution/blob/master/team_minutes/2020-09-08.md)), a short survey was conducted to gather views on potential chat/Q&A tools among the R Foundation and members of the On Ramps team.

As one respondent clarified:

- **R core member** is someone with commit permissions to the R core repository
- **contributor** someone that contributes patches to R that would be applied by an R core member

This initiative seeks to encourage contributors to R core (someone that makes significant contributions might eventually become part of R core, but that is not the main focus here).


## Respondents

- 8 R core members/expert contributors to R core
- 12 intermediate/novice contributors, community organizers or other (R admin/package author)

```{r about}
responses <- responses %>%
  mutate(about = factor(about, 
                        levels = c("I am on R core", 
                                   "I have made frequent/sizable contributions to R core", 
                                   "I have made infrequent/minor contributions to R core",
                                   "I am a novice/potential contributor to R core",
                                   "I am a community organizer",
                                   "Other"),
                        labels = c("R core", "Expert", "Intermediate",
                                   "Novice", "Facilitator", "R admin")),
         expert = if_else(about %in% c("R core", "Expert"), "Expert", "Other"))
ggplot(responses, aes(x = about, fill = about)) +
  geom_bar() + 
  labs(x = NULL, y = "Count", fill = NULL)
```

## Chat Platforms

### Slack

In the initial discussions, there seems to be a preference for Slack, so this was asked about explicitly.

Would they use Slack as a mentoring platform?

 - Experts: 3 yes, 3 maybe, 2 no
 - Other: 11 yes, 1, maybe
 
```{r slack}
responses <- responses %>%
  mutate(slack = factor(slack, c("Yes", "Maybe", "No")))
ggplot(responses, aes(x = slack, fill = about)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count", fill = NULL)
```

### Preferred platform

Respondents were asked to (partially) rank their preferred platform out of Slack, Zulip, Mattermost, IRC and email mailing list. A comparison table of these platforms was given for background information ([chat_platforms.csv](chat_platforms.csv)).

What is the platform they would most prefer to use?

 - Experts: split across Slack, Zulip and email
 - Other: the one that would maybe use Slack prefers email
 
Of the experts that would not use Slack, one would use email, one would not use any platform (not enough time to engage in chat-style communication, only hope would be asynchronous messaging).

Some strongly prefer Zulip since it is FOSS.
 
```{r preferred}
responses <- responses %>%
  mutate(slackalt.1. = factor(slackalt.1., c("Slack", "Zulip", "Email", ""),
                              c("Slack", "Zulip", "Email", "No pref\n/none")))
ggplot(responses, aes(x = slackalt.1., fill = slack)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count",
       fill = "Would use slack?") + 
  scale_fill_viridis_d()
```

### Zulip

Would it be better to use Zulip?

 - Experts: brings maybes on board, but most would not use
 - Other: most would not use
 
Many do not know Zulip and would prefer not to use a new tool (one suggested Discord, but likely that is also new to many).

```{r zulip}
responses <- responses %>%
  mutate(zulip = case_when(
           slackalt.1. == "Zulip" | slackalt.2. == "Zulip" | slackalt.3. == "Zulip" ~ "Yes",
           TRUE ~ "No"
         ),
         zulip = factor(zulip, c("Yes", "No")))
ggplot(responses, aes(x = zulip, fill = slack)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count", fill = "Would use Slack?") + 
  scale_fill_viridis_d()
```

### Email

Would it be better to use Email?

 - Experts: brings most on board, though most of these would/might use Slack
 - Other: better than Zulip, but still lose some intermediate/novice contributors
 
```{r email}
responses <- responses %>%
  mutate(email_rank = case_when(
           slackalt.1. == "Email" ~ 1, 
           slackalt.2. == "Email" ~ 2,
           slackalt.3. == "Email" ~ 3,
           slackalt.4. == "Email" ~ 4,
           TRUE ~ NA_real_
         ),
         email = factor(!is.na(email_rank), c(TRUE, FALSE), c("Yes", "No")),
         email_rank = fct_explicit_na(as.factor(email_rank), na_level = "NA"))
ggplot(responses, aes(x = email, fill = slack)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count", fill = "Would use Slack?") + 
  scale_fill_viridis_d() + 
  scale_y_continuous(breaks = c(3, 6, 9))
```

For most people it is not their first choice and is 3rd or 4th choice among many non-experts.

```{r email_rank}
ggplot(responses, aes(x = email, fill = email_rank)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count", fill = "Choice order") + 
  scale_fill_viridis_d(option = "A") + 
  scale_y_continuous(breaks = c(3, 6, 9))
```

### Proposal

Start Slack group and see if engagement from experts is sufficient to be worth it.

If people are invited to Slack - even if they do not join - they can receive email notifications when people contact them use @mentions, direct messages or @channel. So we could invite interested R core members to a special channel and use @channel when help is needed beyond the people active on Slack - they can then respond by email and the reply will appear in Slack.

## Q&A Platforms

Would people use a Q&A platform?

 - Experts: much the same attitude as towards the chat platform
 - Other: much more uncertain, particularly novice/potential contributors

```{r q-and-a}
responses <- responses %>%
  mutate(QandA = factor(QandA, c("Yes", "Maybe", "No")))
ggplot(responses, aes(x = QandA, fill = about)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count", fill = NULL)
```

### Preferred platform

Respondents were asked to (partially) rank their preferred platform out  of Biostar, Discourse, Scoold and Talkyard. Again, a comparison table was provided ([q_and_a.csv](chat_platforms.csv)).

Clear preference for Discourse. Reasons:

- Only know Discourse
- FOSS
- Email digest
- No separate chat function (assuming we already have a Slack/email group)

```{r preferred-q_and_a}
responses <- responses %>%
  mutate(QandAalt.1. = factor(QandAalt.1., c("Discourse", "Biostar", ""),
                              c("Discourse", "Biostar", "No pref\n/none")))
ggplot(responses, aes(x = QandAalt.1., fill = QandA)) +
  geom_bar() + 
  facet_wrap(~ expert) + 
  labs(x = NULL, y = "Count",
       fill = "Would use any?") + 
  scale_fill_viridis_d()
```
 
Notes:

 - All alternatives (Discourse, Biostar, Scoold, Talkyard) are FOSS, though for Scoold we would like want features in Scoold Pro (e.g. email digest, account suspension, favourite posts, etc) and that is closed source.
 - Only Talkyard has separate chat function which can be disabled.
 
### Proposal

Stick with Slack/email for now, consider setting up Discourse further down the line.



 
 
