---
title: "Installing R from Source on macOS"
output: github_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Prerequisites

Some details and notes are hidden by default - click the black arrowhead to 
expand or collapse these sections.

### Set up installation directories

If you do not have an existing version of R installed on your computer, download 
the installer for the latest release of R from the [CRAN macOS page](https://cran.r-project.org/bin/macosx/) and install this pre-built 
(binary) version with the default settings.

Open Terminal.app to run the shell commands in the instructions below:

1.  Make your account own the R framework folder (created when installing R from 
the binary)
    ```{sh}
sudo chown -R $USER /Library/Frameworks/R.framework/
   ```
2. Make your account own `/opt/R` so that we can install prerequisites for 
building from source with the `install.libs` R function described in the next subsection.
    ```{sh}
sudo chown -R $USER /opt/R
    ```
    <details><summary><i>Details...</i></summary> If this is skipped, you get errors when 
running `install.libs("r-base-dev")`, e.g. "Can't restore time", 
"Can't unlink already-existing object".
    </details>
3.  Open the profile for the Zsh shell 
    ```{sh}
    open /Users/$USER/.zprofile
    ```
    Add one of the following commands to the profile, selecting the one that 
    matches your architecture:
    ```{sh}
# Intel
export PATH="/opt/R/x86_64/bin:\${PATH}"
    ```
    ```{sh}
# Apple Silicon (M1, M2, ...)
export PATH="/opt/R/arm64/bin:\${PATH}"
    ```
    This will add the `/opt/R/*/bin` directory to your `PATH` variable 
    when you restart Terminal.app, so that the prerequisites 
    installed with `install.libs` can be found.
   <details><summary><i>Note if using RStudio terminal, or using Terminal.app on macOS < 10.15 (Catalina)...</i></summary> The code above assumes you are running shell commands in a Zsh shell (the default for Terminal.app on macOS &ge; 10.15). If you are using a bash shell, edit <code>.bash_profile</code> rather than <code>.zprofile</code>. The type of terminal in RStudio can be set in Terminal Options from the terminal or in Global Options.
</details>

### Install prerequisites

Start a new R session. For the following instructions, run the code within R:
	
[//]: # (
check .zprofile okay? 
system2 which svn - need to restart computer or something for it to be recognised once installed?
)

1.  Use `install.libs` from <https://mac.R-project.org> to install commonly required prerequisites
    ```{r}
source("https://mac.R-project.org/bin/install.R")
install.libs("r-base-dev")
    ```
2.  Check if you already have Subversion installed:

    ```{r}
system2("which", "svn")
    ```
    
    This will print a path where Subversion is installed. If nothing is 
    printed, install Subversion:

    ```{r}
install.libs("subversion", dep = FALSE)
    ```
    
    <details><summary><i>Troubleshooting...</i></summary> If this does not work, try installing Subversion via your preferred package manager, e.g. with Homebrew: <code>brew install svn</code>.
    </details>
    
For the remaining instructions, code should be run within Terminal.app

1.  Install GNU Fortran (gfortran) from the binary at 
<https://mac.r-project.org/tools/>. This version is prepared specifically 
for R and will be used in preference to any other version on your system.
2.  From the terminal, install Xcode Command Line Tools
    ```{sh}
    xcode-select --install
    ```
    (If already installed, shows "xcode-select: error: command line tools are
already installed").
3.  Check if you have a current version of MacTeX installed:
    ```{sh}
ls /usr/local/texlive
    ```
    If the `/usr/local/texlive` directory exists, you will see one or more subdirectories named by year, e.g.
    ```
    2022            2023            texmf-local
    ```
    If this check doesn't show a subdirectory for the current year, install MacTex from the binary at <https://tug.org/mactex/mactex-download.html>.
4.  Check if you have XQuartz installed:
    ```{sh}
mdls -name kMDItemVersion /Applications/Utilities/XQuartz.app
    ```
    This will show the version number or `could not find /Applications/Utilities/XQuartz.app`. If missing, install XQuartz from the binary at  <https://www.xquartz.org/>. **Restart your computer** for XQuartz to work!  
5. (Recommended)  If you are not already using a tool such as [rig](https://github.com/r-lib/rig) 
to install and manage multiple R versions, install RSwitch from <https://rud.is/rswitch/>. These third-party tools are not part of the R 
Project, but are recommended here as they provide an easy way to switch R 
versions.

<details><summary><i>Detail on selected prerequisites...</i></summary>
The following prerequisites have been installed by the above instructions:
<ul>
<li> jpeg, libpng, pkgconfig, tiff and zlib-system-stub from r-base-dev. </li>
<li> Objective C compiler from xcode command line tools. </li>
<li> Tcl/Tk and texinfo from standard installation of R binary. </li>
</ul>
The following optional prerequisites have been skipped as they are unlikely 
to be needed by R contributors
<ul>
<li> readline5. Apple's editline is sufficient. Can install with `install.libs("readline5")`. </li>
<li> Cairo/Pango. Can install both with `install.libs("pango")`. </li>
<li> Java. See [Java subsection in R-admin macOS instructions](https://cran.r-project.org/doc/manuals/r-patched/R-admin.html#Java-_0028macOS_0029) to add Java support. You may have installed Java to use a Java-using R package. </li>
</ul>
</details>

## Build R

### Checking out a source code repository

Run the following commands within Terminal.app to check out a local copy of the 
R Subversion (SVN) repository into `TOP_SRCDIR`:
```{sh}
export TOP_SRCDIR="$HOME/svn/R-devel"
svn checkout https://svn.r-project.org/R/trunk/ "$TOP_SRCDIR"
```
This will retrieve the source code for the development version of R (`r-devel`).

<details><summary><i>Alternatively create an RStudio SVN project...</i></summary> You may prefer to create an RStudio project from the SVN repository. Set `https://svn.r-project.org/R/trunk/` as the Repository URL; set `R-devel` as the project directory name and create the project as a subdirectory of `/Users/<username>/svn` (or adapt the definition of `TOP_SRCDIR` to match where you create the project). This will create an SVN pane in RStudio where you can track changes.
</details>

<details><summary><i>Alternatively checkout a fork of the r-svn GitHub repo...</i></summary> You may prefer to use a fork of the <a href="https://github.com/r-devel/r-svn">r-svn GitHub repo</a>. Set the <code>TOP_SRCDIR</code> environment variable to the directory where you checkout your fork of the repository.
</details>

### Building R from a new checkout
    
To build R from a local copy of the SVN repository for the first time, run the 
following commands within Terminal.app: 

1.  Download the source code for the recommended packages:
    ```{sh}
$TOP_SRCDIR/tools/rsync-recommended
    ```

2.  Create the build directory, `BUILDDIR` (you may wish to customise this path):
    ```{sh}
export BUILDDIR="$HOME/build/R-devel"
mkdir -p $BUILDDIR
cd $BUILDDIR
    ```

3.  Define the `LOCAL` environment variable to match your architecture
    ```{sh}
# Intel
export LOCAL=/opt/R/x86_64 
    ```
    ```{sh}
# Apple Silicon (e.g. M1)
export LOCAL=/opt/R/arm64  
    ```
    Create `config.site` within the build directory to set some configuration flags as recommended by the [R-admin manual](https://cran.r-project.org/doc/manuals/r-devel/R-admin.html#Prerequisites).
    ```{sh}
cat << EOF > config.site
CC=clang
OBJC=\$CC
FC="/opt/gfortran/bin/gfortran -mtune=native"
CPPFLAGS='-isystem $LOCAL/include'
CXX=clang++
PKG_CONFIG_PATH=$LOCAL/lib/pkgconfig:/usr/lib/pkgconfig
EOF
    ```
    If you are on Apple Silicon, also run the following to add some more configuration flags
    ```{sh}
cat << EOF >> config.site
CFLAGS="-falign-functions=8 -g -O0"
FFLAGS="-g -O2 -mmacosx-version-min=11.0"
FCFLAGS="-g -O2 -mmacosx-version-min=11.0"
LDFLAGS="-L$LOCAL/lib -L/opt/gfortran/lib"
CPPFLAGS="-isystem $LOCAL/include -I$LOCAL/include"
EOF
    ```
    <details><summary><i>Details...</i></summary> Some modifications to the recommendations in R-admin: `-O0` to enable debugging symbols and disable compiler optimisations for better debugging experience; `-mmacos-version-min` corrected [?] to `-mmacosx-version-min`; `LDFLAGS="-L$LOCAL/lib -L/opt/gfortran/lib"` added so that liblzma (in `$LOCAL/lib`) and gfortran libraries can be found. CPPFLAGS modified for Apple Silicon to link to the headers for liblzma.
    </details>
    <details><summary><i>Extra step if you are using the r-svn GitHub mirror...</i></summary>You will need to update the Makefile template to infer the SVN revision number from the git mirror. Run the following line of code to replace an `svn` command in the template with a shell script that will infer the SVN revision number: <br>
    <code>
    sed -i.bak \"s|\\\\$(GIT) svn info|\$TOP_SRCDIR/.github/scripts/svn-info.sh|\" \"\$TOP_SRCDIR/Makefile.in\"
    </code>
    </details>

4. Configure the R installation with `--enable-R-framework` to prepare for installation as an App and `--disable-java` to skip configuring Java, which may not be installed. Use `FW_VERSION` to set the version name as "R-devel":
    ```{sh}
$TOP_SRCDIR/configure --enable-R-framework --disable-java FW_VERSION=R-devel
    ```
    <details><summary><i>Multiple versions of R-devel...</i></summary> If you want to build multiple versions of R-devel, customize `FW_VERSION` to give a unique name to each build. Be aware that if you install R-devel from an Intel Mac binary, e.g., from <url>https://mac.r-project.org/</url> or using rig, this will use the default name "Major.Minor", using the major and minor numbers from the R version number (e.g. 4.4 for R 4.4.0 Under development version).
    </details>
    
[//]: # (
The configure options in [R-admin manual](https://cran.r-project.org/doc/manuals/r-devel/R-admin.html#Prerequisites) to define the location of X11 and tcltk libraries do not seem to be necessary.
)
    
5.  Build R :
    ```{sh}
make
    ```
6.  Check that R works as expected:
    ```{sh}
make check
    ```
    `make` will exit with an error if there are any problems. 
    You can start the built version of by running the command `bin/R` from 
    the build directory, but we recommending installing as an App, as 
    described next.
    <details><summary><i>Note if you are using RStudio terminal...</i></summary> `make check` will fail unless you run `unset R_HOME` first
    </details>
7.  Install the built version of R and reset permissions on the 
R framework folder to include the new directories.
    ```{sh}
make install
sudo chown -R $USER /Library/Frameworks/R.framework/
    ```
    The built version of R will now be your default version of R, which you can start by typing `R` in the terminal, or restarting the R session in RStudio, etc. To switch versions, use RSwitch or rig, then restart your R session.

## Re-build R

After making changes to the source files in `$TOP_SRCDIR` (via `svn update` or making your own modifications), 
you can re-build R incorporating the changed components by switching to the 
build directory in the terminal and running make:

```{sh}
export BUILDDIR="$HOME/build/R-devel"
cd $BUILDDIR
make
```

If you are building a version with your own changes, you may wish to run 
`make check` to check you haven't broken any tests.

If you followed the recommended steps to install R as an App, you can install 
the re-built version as follows:

```{sh}
make install
```

