---
title: "Installing R from Source on macOS"
output: github_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Prerequisites

Some details and notes are hidden by default - click the black arrowhead to 
expand or collapse these sections.

### Set up installation directories

If you do not have an existing version of R installed on your computer, download 
the installer for the latest release of R from the [CRAN macOS page](https://cran.r-project.org/bin/macosx/) and install this pre-built 
(binary) version with the default settings.

Open Terminal.app to run the shell commands in the instructions below:

1.  Make your account own the R framework folder (created when installing R from 
    the binary). Make sure you re-run this command any time you re-install R.
    ```{sh}
    sudo chown -R $USER /Library/Frameworks/R.framework
    ```
2.  Make your account own `/opt/R` so that we can install prerequisites for 
    building from source with the `install.libs` R function described in the next subsection.
    ```{sh}
    sudo chown -R $USER /opt/R
    ```
    <details><summary><i>Details...</i></summary> If this is skipped, you get errors when 
running `install.libs("r-base-dev")`, e.g. "Can't restore time", 
"Can't unlink already-existing object".
    </details>
3.  Run the following command to add the necessary tools to your `PATH`:
    ```{sh}
    export PATH="/opt/R/$(uname -m)/bin:/opt/gfortran/bin:${PATH}"
    ```
    This will add the `/opt/R/*/bin` directory and GNU Fortran to your `PATH` variable
    which enables your shell to find those tools. You have to do this whenever you open a new shell.
    
    <details><summary><i>You can make it persistent by editing your profile...</i></summary>
    You can add the above line to your profile so it will be added automatically,
    but this requires you to edit `~/.profile`, `~/.zprofile` or `~/.zshrc` depending on
    your version of macOS and your local settings. Typically, if any of the above files
    exists, you add it there, otherwise you have use the file that corresponds to your
    shell (`~/.zprofile` for zsh and `~/.profile` for bash).</details>

4.  Make sure that you rename `~/.R/Makevars` file if it exists with
    ```{sh}
    [[ -e ~/.R/Makevars ]] && mv ~/.R/Makevars ~/.R/Makevars.old
    ```
	<details><summary><i>Details...</i></summary>Other toolchains may have created custom compiler flags that
	are incomaptible with "native" R build so we want to make sure they are not
	in the way.</details>


### Install prerequisites

Start a new R session in the Terminal.app after setting the `PATH` as above by typing `R`. For the following instructions, run the code within R:
	

1.  Use `install.libs` from <https://mac.R-project.org> to install commonly required prerequisites
    ```{r}
    source("https://mac.R-project.org/bin/install.R")
    install.libs("r-base-dev")
    ```

2.  Check if you already have Subversion installed:
    
    ```{r}
    Sys.which("svn")
    ```
    
    This will print a path where Subversion is installed. If only empty string is 
    printed, install Subversion:
    
    ```{r}
    install.libs("subversion", dep = FALSE)
    ```
     
    <details><summary><i>Troubleshooting...</i></summary>Please make sure you have restarted-your Terminal.app after changing the profile so your `PATH` changes are in effect. Check with `echo $PATH` to make sure it starts with `/opt/R..`. If this does not work, try installing Subversion via your preferred package manager, e.g. with Homebrew: `brew install svn`, but some Homebrew packages cause conflicts so only use as last resort. 
    </details>
    
For the remaining instructions, code should be run within Terminal.app

1.  Install GNU Fortran (gfortran) from the binary at 
	<https://mac.r-project.org/tools/>. This version is prepared specifically 
	for R and will be used in preference to any other version on your system.

2.  From the terminal, install Xcode Command Line Tools
    ```{sh}
    xcode-select --install
    ```
    (If already installed, shows "xcode-select: error: command line tools are
	already installed").

3.  Check if you have LaTeX installed:
    ```{sh}
    pdflatex --version | head -n1
    ```
    It should show something like
    ```
    pdfTeX 3.141592653-2.6-1.40.25 (TeX Live 2023)
    ```
    If it shows an error or the year is too old, install MacTex from the binary at <https://tug.org/mactex/mactex-download.html> (or install TinyTex if you prefer).

4.  Check if you have XQuartz installed:
    ```{sh}
    mdls -name kMDItemVersion /Applications/Utilities/XQuartz.app
    ```
    This will show the version number or `could not find /Applications/Utilities/XQuartz.app`. If missing, install XQuartz from the binary at  <https://www.xquartz.org/>. **Restart your computer** for XQuartz to work!  

5. (Recommended)  If you are not already using a tool such as [rig](https://github.com/r-lib/rig) 
   to install and manage multiple R versions, install RSwitch from <https://rud.is/rswitch/>. These third-party tools are not part of the R 
   Project, but are recommended here as they provide an easy way to switch R versions.

<details><summary><i>Detail on selected prerequisites...</i></summary>
The following prerequisites have been installed by the above instructions:
<ul>
<li> jpeg, libpng, pkgconfig, tiff and zlib-system-stub from r-base-dev. </li>
<li> Objective C compiler from xcode command line tools. </li>
<li> Tcl/Tk and texinfo from standard installation of R binary. </li>
</ul>
The following optional prerequisites have been skipped as they are unlikely 
to be needed by R contributors
<ul>
<li> readline5. Apple's editline is sufficient. Can install with `install.libs("readline5")`. </li>
<li> Cairo/Pango. Can install both with `install.libs("pango")`. </li>
<li> Java. See [Java subsection in R-admin macOS instructions](https://cran.r-project.org/doc/manuals/r-patched/R-admin.html#Java-_0028macOS_0029) to add Java support. You may have installed Java to use a Java-using R package. </li>
</ul>
</details>

## Build R

### Checking out a source code repository

Run the following commands within Terminal.app to check out a local copy of the 
R Subversion (SVN) repository into `R-devel` subdirectory.
```{sh}
svn checkout https://svn.r-project.org/R/trunk/ R-devel
```
This will retrieve the source code for the development version of R (`r-devel`).

<details><summary><i>Alternatively create an RStudio SVN project...</i></summary> You may prefer to create an RStudio project from the SVN repository. Set `https://svn.r-project.org/R/trunk/` as the Repository URL; set `R-devel` as the project directory name and create the project as a subdirectory. This will create an SVN pane in RStudio where you can track changes.
</details>

<details><summary><i>Alternatively clone a fork of the r-svn GitHub repo, however, that requires `git-svn` to be installed manually.</i></summary> You may prefer to use a fork of the <a href="https://github.com/r-devel/r-svn">r-svn GitHub repo</a>. If you do, you must create `SVNINFO` in the `R-devel` directory with `git svn info > SVNINFO`, however, this requires `git-svn` which is not installed by default on macOS.
</details>

### Building R from a new checkout

To build R from a local copy of the SVN repository for the first time, run the 
following commands within Terminal.app: 

1.  Download the source code for the recommended packages:
    ```{sh}
	( cd R-devel && tools/rsync-recommended )
    ```
	(If your network blocks `rsync` use `tools/fetch-recommended` instead)

2.  Create the `build` subdirectory where R will be built and make it your current directory:
    ```{sh}
	mkdir build
	cd build
    ```

3.  Define the `LOCAL` environment variable to match your architecture so you don't have to type the same path repeatedly
    ```{sh}
    export LOCAL=/opt/R/$(uname -m)
    ```

4. Configure the R installation with:
   
   ```{sh}
   ../R-devel/configure --prefix=$LOCAL --enable-R-framework --enable-R-shlib \
      CPPFLAGS=-I$LOCAL/include LDFLAGS=-L$LOCAL/lib CFLAGS='-Wall -g -O0' \
      --x-includes=/opt/X11/include --x-libraries=/opt/X11/lib \
      FC=/opt/gfortran/bin/gfortran
   ```
   <details><summary><i>Multiple versions of R-devel...</i></summary> If you want to build multiple versions of R-devel, add `FW_VERSION=name` to give a unique `name` to each build. Be aware that if you install R-devel from an Intel Mac binary, e.g., from <url>https://mac.r-project.org/</url> or using rig, this will use the default name "Major.Minor", using the major and minor numbers from the R version number (e.g. 4.4 for R 4.4.0 Under development version).
    </details>
    
[//]: # (
The configure options in [R-admin manual](https://cran.r-project.org/doc/manuals/r-devel/R-admin.html#Prerequisites) to define the location of X11 and tcltk libraries do not seem to be necessary.
)
    
5. Build R :
   ```{sh}
   make -j8
   ```
   (The `-j8` instructs `make` to use 8 parallel jobs. You can either increase the number corresponding to you machine's capabilities, or drop it when debugging since it can be harder to find the error among parallel jobs).

6. Check that R works as expected:
   ```{sh}
   make check
   ```
   `make` will exit with an error if there are any problems. 
   You can start the built version of by running the command `bin/R` from 
   the build directory.
   <details><summary><i>Note if you are using RStudio terminal...</i></summary> `make check` will fail unless you run `unset R_HOME` first
   </details>
7. Install the built version of R
   ```{sh}
   make install
   ```
   The built version of R will now be your default version of R, which you can start by typing `R` in the terminal, or restarting the R session in RStudio, etc. To switch versions, use RSwitch or rig, then restart your R session.

## Re-build R

After making changes to the source files in `R-devel` (via `svn update` or making your own modifications), 
you can re-build R incorporating the changed components by switching to the 
build directory in the terminal and running `make` again.

If you are building a version with your own changes, you may wish to run 
`make check` to check you haven't broken any tests.

If you followed the recommended steps to install R as an App, you can install 
the re-built version with `make install` as before.

NOTE: Sometimes re-building signed binaries can cause problems which manifest themselves as `Killed: 9` in the error. If that is the case, the most reliable way to resolve it is to remove the `build` directory with `rm -rf build`, and repeat steps from 2. on.
