# Evaluating the status of translations in R

The script `translations_status.R` produces datasets regarding message translations from a local clone of https://github.com/r-devel/r-svn, 
a mirror of the SVN repository containing the source code for R. 
The script is designed to be run interactively and requires the file path of the `r-svn` repo and an output directory to be specified.
The datasets are saved in the output directory with the short SHA and date of the git commit on which they were based as a suffix.

## Languages CSV

A CSV with the variables:

- `package`: the name of a package containing messages to be translated
- `po_file`: the name of `.po` files in the package sources
- `component`: the "component" of the package the PO file relates to, either "C", "R", "RGui" (the latter is only in the base package)
- `code`: the ISO 639 code for the language the PO file relates to
- `variant`: the variant of the language, if applicable (e.g. BR for pt-BR)
- `language`: the name of the language in English
- `country`: the country corresponding to the variant

## Metadata CSV

A CSV with the variables:

- `package`: the name of a package containing messages to be translated
- `po_file`: the name of `.po` files in the package sources
- `r_version`: the name of the R version to PO file relates do (does not always match `pot_creation_date`)
- `bug_reports`: where to report bugs related to this PO file
- `pot_creation_date`: the date the PO template file was created (when messages last updated), YYYY-MM-DD format
- `po_revision_date`: the date the PO file was revised, YYYY-MM-DD format
- `last_translator`: the name and email of the last translator
- `team`: the name and/or email of the translation team

## Message status CSV

A CSV with the variables:

- `package`: the name of a package containing messages to be translated
- `po_file`: the name of `.po` files in the package sources
- `message`: a message in the PO file
- `translated`: a logical value indicating if the message has been translated
- `fuzzy`: a logical value indicating if the translation is flagged as "fuzzy", i.e. a fuzzy match of an old translation to a message that has had a minor update

## Example plot

```{r, message = FALSE}
library(dplyr)
library(forcats)
library(ggplot2)
library(readr)
```

```{r, message = FALSE}
message_status <- read_csv("message_status_4168b6f_2022-04-25.csv")
languages <- read_csv("languages_4168b6f_2022-04-25.csv")
```

First a little tidying to create short names for the dialects
```{r}
plot_dat <- left_join(message_status, languages) |>
    mutate(dialect = ifelse(is.na(variant),
                            gsub("([^;]+).*", "\\1", language),
                            paste(language, variant, sep = "_")))
```                            
Then we can plot the counts of translated messages
```{r}
ggplot(filter(plot_dat, translated),
       aes(x = fct_infreq(dialect))) +
    geom_bar(stat = "count", fill = "steelblue") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
          legend.position = "none") +
    labs(x = NULL, y = "# translated messages",
         subtitle = "Translated messages in base and default packages")
```
There is one message translation in standard English - this is to add information about the locale to the startup message in R.
